Implementation of a lattice Boltzmann method for immiscible multiphase flow simulations using the level set method

### 2015-11-11 Paper/paper.tex
Kleine Änderungen am Paper
Wir sollten vlt auch noch auf doi:10.1016/j.camwa.2009.02.005
referenzieren (das zweie Thoemmes paper)

The evolution of the flow fields within each fluid can be expressed with the incompressible Navier-Stokes equations:
\begin{equation}
	\nabla \cdot  \vec u_i = 0 \text{  in } \Omega_i \text{,}
\end{equation}
\begin{equation}
\frac{\partial \vec u_i}{\partial t} + \left(\vec  u_i\cdot \nabla \right) \vec u_i
=
-\frac{1}{\varrho_i}\nabla p_i + \nu_i \nabla^2 \vec u_i \text{  in } \Omega_i \text{.}
\end{equation}
Here, $u_i$ is the velocity, $p_i$ the pressure, $v_i$ the kinematic viscosity and $\varrho_i$ the mass density.

Furthermore, we need boundary conditions and initial conditions. For our domain-boundaries and the interior we can take for example
\begin{equation}
	u_i(x,t) = 0 \text{, on } \delta\Omega_i \backslash \Gamma \text{,}
\end{equation}
\begin{equation}
	u_i(x,0) = u_{ini,i}(x) \text{, in } \Omega_i \text{.}
\end{equation}
For the interface $\Gamma$ we have the influence of the surface tension and we assume continuity of the velocities within the fluids. Therefore, we get the following jump conditions:
\begin{equation}
	[u] = 0 \text{,  } [T]n = 2 \sigma \kappa n \text{.}
	\label{jumpconditions}
\end{equation}
Here, T is the stress tensor, $\sigma$ the surface tension, and $\kappa$ the curvature of the interface with respect to its normal $n$. Brackets denote the jump of a quantity at the interface: 
$[q](x) = \lim_{\epsilon \rightarrow 0} (q(x - \epsilon n) - (q(x + \epsilon n)) \text{ with } x \in \Gamma$ % I also turned this around, according to [p].
. In Fig. \ref{intro_phases} this would be the jump from $\Omega_1$ to $\Omega_2$. For the other direction the sign changes.

@@

The method we are presenting uses a hybrid lattice Boltzmann level set method. Thereby the surface motion is calculated by the level set method. The input data for the level set method is given by the lattice Boltzmann method, which is calculated independently for each fluid. This method has the advantage that it has a high accuracy but requires only little memory.

@@

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Navier-Stokes equations, we use the lattice Boltzmann method. This method describes the evolution of the particle density $f(\vec{x},\vec{v},t)$ in phase space, with a $(\vec{x},\vec{v})$ as the phase space variables and $t$ as time. The LBM algorithm discretizes with a regular grid in space and with restricted number of velocities adapted to this grid \cite{LBM1,LBM2,LBM3,LBM4}. We use the D2Q9 model in 2D, which has 9 velocity vectors on a plane grid with unit spacing including one zero velocity (Fig. \ref{lbmc_i}). 
\begin{figure}
	\hfill\includegraphics[width=6cm,natwidth=403,natheight=409]{LBMc_i.png}\hspace*{\fill}
	\caption{The vectors of one cell in LBM}
	\label{lbmc_i}
\end{figure}
The vectors are given by the columns of the matrix
\begin{equation}
	c = 
	\begin{bmatrix}
	0 & 0 & 1 & 1 &  1 &  0 & -1 & -1 & -1 \\
	0 & 1 & 1 & 0 & -1 & -1 & -1 &  0 &  1
	\end{bmatrix}
	\text{.}
\end{equation}
The corresponding particle distributions are denoted by $f_i(\vec{x},t) = f(\vec{x},\vec{c_i},t)$. We get the density and velocity from the particle distribution functions with:
\begin{equation}
	\rho(\vec{x},t) = \sum_{i=0}^{8} f_i(\vec{x},t) \text{, } 
	\vec{u} = \sum_{i=0}^{8} f_i(\vec{x},t)\vec{c_i} \text{.}
\end{equation}
For the collision step we use the Bhatnagar-Gross-Krook (BGK) approximation \cite{BGK}
\begin{equation}
	f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x},t) - \frac1\tau(f_i-f_i^{eq}) + G_i \text{.}
\end{equation}
The parameter $\tau$ is the relaxation parameter for the BGK collision operator and controls the kinematic viscosity $\nu = \frac16(2\tau - 1)$. $G_i$ is an additional force like gravity or a magnetic field. In our case we set $G_i = 0$. Furthermore, we use the equilibrium distribution
\begin{equation}
	f_i^{eq}(\rho,\vec{u}) = f_i^*(\rho + 3\vec{c_i} \cdot \vec{u} + \frac92(\vec{c_i} \cdot \vec{u})^2 - \frac32{|\vec{u}|}^2)
\end{equation}
with the corresponding D2Q9 weight factors
$$
f_i^* = \begin{cases} \frac49 \text{,} & i = 0 \\ \frac19 \text{,} & i = 1,3,5,7 \\ \frac{1}{36} \text{,} &  i = 2,4,6,8 \end{cases} \text{.}
$$

The LBM algorithm consists out of the following steps:
\begin{itemize}
	\item[1. ] Collision step: $f_i^+ = f_i - \frac1\tau(f_i - f_i ^{eq}) + G_i$
	\item[2. ] Propagation step: $f_i(\vec{x}+\vec{c_i},t+1) = f_i^+(x,t)$, for interior nodes
	\item[3. ] Boundary conditions: $f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x}+\vec{c_i},t)$, if $\vec{x} \notin \Omega$
\end{itemize}

\subsection{The Level Set Method}
We deal with the movement of the interface between the two fluids by using the level set method. The method captures the surface that represents the interface with a continuous signed distance function which has the interface as the zero level set.

Let $\Gamma_t$ be our interface at time t. Within the level set method, $\Gamma_t$ is the zero iso-surface of the level set function $\varphi$. Thus for $x \in \Gamma_t$ we have $\varphi(x,t) = 0$, in direction of the surface normal we get $\varphi > 0$, and in the other direction we get $\varphi < 0$. For this level set function we get the level set equation of Osher and Sethian \cite{OsherSethian} for the evolution of free surfaces
\begin{equation}
	\varphi_t + v \cdot \nabla \varphi = 0 \text{,}
\end{equation}
with the velocity $v = v(x(t)) = \dot x(t)$ for $x(t) \in \Gamma_t$.

@@

For the coupling, we create a new boundary condition that is based on a simple bounce back type Dirichlet condition. For example, in Fig. \ref{coupling} for a point $x_2 \in \Omega_2$ that is next to the boundary we set

@@

	f_i(x_2,t+1) = f_{i*}^+(x_2,t) + 6hf_i^*c_i \cdot \tilde{u} + R_i \text{.}

@@

The additional term $R_i$ ensures the jump conditions of the normal stress and corrects the error terms that are caused by the bounce back treatment. $R_i$ is given as:

@@

	R_i = 6 h^2 f_i^* \Lambda_i : A \text{ with } A = -q(1-q)[S] - (q - \frac12)S^{(2)} \text{.}
We can rewrite the formula in the form
	R_i = 6h^2f_i^* \cdot (-q(1-q)\Lambda_i:[S] - (q-\frac12)\Lambda_i:S_2 \text{,}

@@

\Lambda_i = c_i \otimes c_i - \frac13 {|c_i|}^2 \mathbf{I} \text{,}

@@

@@

	S_k = -\frac{3}{2\tau h^2} \sum_i c_i \otimes c_i(f_i-f_i^{eq})(t,x_k) \text{.}

@@

The tensor product is $a \otimes b = ab^T$ and the double contraction $A:B = \mathrm{trace}(AB^T)$. Furthermore, we have

@@

	\Lambda_i:[S] = ([S]:n \otimes n)((n \cdot c_i)^2 - {|c_i|}^2/3) + 2([S]:n \otimes t_k)(n \cdot c_i)(t_k \cdot c_i) \text{,}

@@

with the terms

@@

@@

	[S]:n \otimes t_k = - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes t_k \text{.}

@@

Here $n$ is the normal of the interface, $t$ the tangential, $\bar{S} = (S_1 + S_2)/2$ the averaged symmetric velocity gradient, $\kappa$ the curvature, and $\mu$ the dynamic viscosity, with $[\mu] = \mu_2 - \mu_1$ as its jump.
The pressure jump is formulated by

@@

	[p] \approx \frac{1}{3 h^2} (\rho(x_2,t)\varrho_2 - \rho(x_1,t)\varrho_1) \text{.}

@@

@@

	\tilde{\rho} = 3\rho(x+c_i,t+1) - 3\rho(x+2c_i,t+1) + \rho(x+3c_i,t+1) \text{.}

@@

Figure \ref{refill} shows the cells we use for the interpolation. In \cite{Thoemmes} the velocity is interpolated by using the velocity of the interface. The Ian-Mitchell toolbox does not provide a function to access this value. Therefore, we interpolate the velocity in the same way as we did for the density:

@@

\tilde{u} = 3u(x+c_i,t+1) - 3u(x+2c_i,t+1) + u(x+3c_i,t+1) \text{.}

@@

We can now calculate the new equilibrium $f_i^{eq}(\tilde{\rho},\tilde{u})$ and add it to the non-equilibrium part that we copy from the direct neighbor:

@@

We have to repeat the procedure for other fluid.
	\hfill\includegraphics[width=8cm,natwidth=632,natheight=562]{dataexchange.png}\hspace*{\fill}
	\caption{Data exchange between LBM and level set method}
	\label{dataexchange}

Finally, we can couple the two methods in the following way:
\begin{itemize}
	\item[1.] Create the initial interface $\Gamma_0$
	\item[2.] Run the level set code once to create the surface description for LBM
	\item[3.] Run the LBM using the data from the level set for a prescribed number of steps
	\item[4.] Pass the current velocity field to the level set code and run it for the same number of steps (see Fig. \ref{dataexchange})
	\item[5.] Repeat step 3 and 4 until the end of the simulation.
\end{itemize}

@@

\bibitem{GRcg2} {\sc A.K. Gunstensen, D.H. Rothman}, {\em Macroscopic modeling of immiscible flows in three dimensions by a lattice Boltzman method}, Europhys. Lett. 18 (2) (1998) 157-161

@@

% Richtige Abkürzung fehlt hier!!!
\bibitem{Koerner} {\sc C. K\"orner, M. Thies, T. Hofmann, N. Th\"urey, U. R\"ude}, {\em Lattice Boltzmann Model for Free Surface Flow for Modeling Foaming}, J. Statistical Physics Vol. 121 (2005) 179-196

@@

\bibitem{LBM3} {\sc Sauro Succi}, {\em The Lattice Boltzmann Equation for Fluid Dynamics and Beyond}, Clarendon Press, Oxford, 2001

@@

>>>>

The evolution of the flow fields within each fluid can be expressed with the incompressible Navier-Stokes equations:
\begin{equation}
	\nabla \cdot  \vec u_i = 0 \text{  in } \Omega_i \text{,}
\end{equation}
\begin{equation}
\frac{\partial \vec u_i}{\partial t} + \left(\vec  u_i\cdot \nabla \right) \vec u_i
=
-\frac{1}{\varrho_i}\nabla p_i + \nu_i \nabla^2 \vec u_i \text{  in } \Omega_i \text{.}
\end{equation}
Here, $u_i$ is the velocity, $p_i$ the pressure, $\nu_i$ the kinematic viscosity and $\varrho_i$ the mass density.

Furthermore, we need boundary conditions and initial conditions. For our domain-boundaries and the interior we can take for example
\begin{equation}
	\vec u_i(x,t) = 0 \text{, on } \delta\Omega_i \backslash \Gamma \text{,}
\end{equation}
\begin{equation}
	\vec u_i(x,0) = \vec u_{ini,i}(x) \text{, in } \Omega_i \text{.}
\end{equation}
For the interface $\Gamma$ we have the influence of the surface tension and a no-slip condition, i.e. continuity of the velocities between the fluid phases. Therefore, we get the following jump conditions:
\begin{equation}
	[\vec u] = 0 \text{,  } [T] \cdot \vec n = 2 \sigma \kappa \vec n \text{.}
	\label{jumpconditions}
\end{equation}
Here, T is the shear stress tensor, $\sigma$ the surface tension, and $\kappa$ the curvature of the interface with respect to its outer normal $\vec n$. Brackets denote the jump of a quantity at the interface: 
$[q](\vec x) = \lim_{\epsilon \rightarrow 0} (q(\vec x - \epsilon \vec n) - (q(\vec x + \epsilon \vec n)) \text{ with } \vec x \in \Gamma$ % I also turned this around, according to [p].
. In Fig. \ref{intro_phases} this would be the jump from $\Omega_1$ to $\Omega_2$. For the other direction the sign changes.

@@

The method we are presenting uses a hybrid lattice Boltzmann level set method. Thereby the interface motion is calculated by the level set method. The input data for the level set method is given by the lattice Boltzmann method, which is calculated independently for each fluid. This method has the advantage that it allows for big ranges of density and viscosity ratios between the fluid phases 
and has a high accuracy but requires only little memory %TODO das ist zu unscharf, glaube ich
.

@@

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Navier-Stokes equations, we use the lattice Boltzmann method (\cite{Succi}). This method describes the evolution of the particle density $f(\vec{x},\vec{v},t)$ in phase space, with a $(\vec{x},\vec{v})$ as the phase space variables and $t$ as time. The LBM algorithm discretizes with a regular grid in space and with restricted number of velocities adapted to this grid \cite{LBM1,LBM2,LBM3,LBM4}. We use the D2Q9 model in 2D, which has 9 velocity vectors on a plane grid with unit spacing including one zero velocity (Fig. \ref{lbmc_i}). 
\begin{figure}
	\hfill\includegraphics[width=6cm,natwidth=403,natheight=409]{LBMc_i.png}\hspace*{\fill}
	\caption{The vectors of one cell in LBM}
	\label{lbmc_i}
\end{figure}
The vectors are given by the columns of the matrix
\begin{equation}
	c = 
	\begin{bmatrix}
	0 & 0 & 1 & 1 &  1 &  0 & -1 & -1 & -1 \\
	0 & 1 & 1 & 0 & -1 & -1 & -1 &  0 &  1
	\end{bmatrix}
	\text{.}
\end{equation}
The corresponding particle distributions are denoted by $f_i(\vec{x},t) = f(\vec{x},\vec{c_i},t)$. We get the density and velocity from the particle distribution functions with:
\begin{equation}
	\rho(\vec{x},t) = \sum_{i=1}^{9} f_i(\vec{x},t) \text{, } 
	\vec{u} = \sum_{i=1}^{9} f_i(\vec{x},t)\vec{c_i} \text{.}
\end{equation}
For the collision step we use the Bhatnagar-Gross-Krook (BGK) approximation \cite{BGK}
\begin{equation}
	f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x},t) - \frac1\tau(f_i-f_i^{eq}) + G_i \text{.}
\end{equation}
The parameter $\tau$ is the relaxation parameter for the BGK collision operator and controls the kinematic viscosity in lattice units $\nu = \frac16(2\tau - 1)$. Note, that we assumed the cell spacing $\Delta x$ and the time step $\Delta t$ to be unity here. The lattice speed of sound equals to $c_s =\frac{1}{\sqrt{3}}$. $G_i$ is an additional force like gravity. In our case we set $G_i = 0$. Furthermore, we use the equilibrium distribution
\begin{equation}
	f_i^{eq}(\rho,\vec{u}) = f_i^*(\rho + 3\vec{c_i} \cdot \vec{u} + \frac92(\vec{c_i} \cdot \vec{u})^2 - \frac32 \vec u\cdot \vec u)
\end{equation}
with the corresponding D2Q9 weight factors
$$
f_i^* = \begin{cases} \frac49 \text{,} & i = 0 \\ \frac19 \text{,} & i = 1,3,5,7 \\ \frac{1}{36} \text{,} &  i = 2,4,6,8 \end{cases} \text{.}
$$

The LBM algorithm consists out of the following steps:
\begin{itemize}
	\item[1. ] Collision step: $f_i^+ = f_i - \frac1\tau(f_i - f_i ^{eq}) + G_i$
	\item[2. ] Propagation step: $f_i(\vec{x}+\vec{c_i},t+1) = f_i^+(x,t)$, for interior nodes
	\item[3. ] Boundary conditions: $f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x}+\vec{c_i},t)$, if $\vec{x} \notin \Omega$
\end{itemize}
% TODO Propagation kurz im Text erklären. vlt Bounce-Back boundary conditions erklaeren?

\subsection{The Level Set Method}
We deal with the movement of the interface between the two fluids by using the level set method. The method captures the surface that represents the interface with a continuous signed distance function, where the interface is implicitly given as the zero level set.

Let $\Gamma_t$ be our interface at time t. Within the level set method, $\Gamma_t$ is the zero iso-surface of the level set function $\varphi$. Thus for $\vec x \in \Gamma_t$ we have $\varphi(\vec x,t) = 0$, in direction of the surface normal we get $\varphi > 0$, and in the other direction we get $\varphi < 0$. For this level set function we get the level set equation of Osher and Sethian \cite{OsherSethian} for the evolution of free surfaces
\begin{equation}
  \frac{\partial \varphi}{\partial t} + \vec v \cdot \nabla \varphi = 0 \text{,}
\end{equation}
with the velocity $\vec v = \vec u(\vec x(t))$ for $\vec x(t) \in \Gamma_t$, which we obtain from the lattice Boltzmann method.

@@

For the coupling, we create a new boundary condition that is based on a bounce back type condition, that prescribes the velocity as  Dirichlet value. For example, in Fig. \ref{coupling} for a point $x_2 \in \Omega_2$ that is next to the phase interface we set

@@

	f_i(x_2,t+1) = f_{i*}^+(x_2,t) + 6f_i^*c_i \cdot \tilde{u} + R_i \text{.}

@@

The additional term $R_i$ ensures the jump conditions of the normal stress and corrects the error terms that is introduced by the first order accurate bounce back treatment. $R_i$ is given as:

@@

	R_i = 6 f_i^* \Lambda_i : A \text{ with } A = -q(1-q)[S] - (q - \frac12)S^{(2)} \text{, }
  \label{add_term2}

@@

  \Lambda_i = c_i \otimes c_i - \frac12 (c_i \cdot c_i) \mathbf{I} \text{. }

@@

That is, an orthogonal projection of the column vectors of the (symmetric) deviatoric shear rate jump tensor on the link direction weighted with the interface distance and a compensation term for the interpolation error that arises, when the interface is not on the exact middle between two LBM grid cells.
The tensor product is $a \otimes b = ab^T$ and the double contraction $A:B = \mathrm{trace}(AB^T)$. We can rewrite formula (\ref{add_term2}) in the form

@@

  R_i = 6f_i^* (-q)(1-q)\Lambda_i:[S] - (q-\frac12)\Lambda_i:S_2 \text{,}

@@

and use the following first-order estimate to obtain the shear rate from the non-equilibrium parts of the distributions (see \cite{Krueger})

@@

  S_k = -\frac{3}{2\tau } \sum_{i=1}^9 c_i \otimes c_i(f_i-f_i^{eq})(t,x_k) \text{.}

@@

Furthermore, we have
  \Lambda_i:[S] = ([S]:n \otimes n)((n \cdot c_i)^2 - \frac12 (c_i \cdot c_i)) + 2([S]:n \otimes t)(n \cdot c_i)(t \cdot c_i) \text{,}
which is obtained by a coordinate transformation onto a basis $(n, t)$ of normal and tangential vector to the interface. From the fluid mechanical jump condition (\ref{jumpconditions}) and the relation $[\mu S] = \bar{\mu}[S] + [\mu]\bar{S} $ it can be seen:

@@

and

@@

	[S]:n \otimes t = - \frac{[\mu]}{\bar{\mu}} \bar{S} : n \otimes t \text{.}

@@

Here $\mu_k = \varrho_k \nu_k$ are the dynamic viscosities, with $\bar{\mu} = (\mu_1 + \mu_2)/2$ and $\bar{S} = (S_1 + S_2)/2$ as averaged quantities, $\kappa$ the interface curvature, and the visocity jump  $[\mu] = \mu_2 - \mu_1$.
The pressure jump is approximated by

@@

	[p] \approx \frac{1}{3} (\rho(x_2,t)\varrho_2 - \rho(x_1,t)\varrho_1) \text{.}

@@

and with that, all variables for the boundary conditions are well defined.

@@

  \tilde{\rho}(x) = 3\rho(x+c_i,t+1) - 3\rho(x+2c_i,t+1) + \rho(x+3c_i,t+1) \text{.}

@@

Figure \ref{refill} shows the cells we use for the interpolation. In \cite{Thoemmes} the velocity is interpolated by using the velocity of the interface. The Ian-Mitchell toolbox does not provide a function to access this value. %TODO das ist misverständlich.
Therefore, we interpolate the velocity in the same way as we did for the density:

@@

  \tilde{u}(x) = 3u(x+c_i,t+1) - 3u(x+2c_i,t+1) + u(x+3c_i,t+1) \text{.}

@@

We can now calculate the new equilibrium part of the particle densities in $x$, $f_i^{eq}(\tilde{\rho},\tilde{u})$ and add them to the non-equilibrium part which we copy from a nearest neighbor:

@@

This completes the refill algorithm.

\begin{figure}
	\hfill\includegraphics[width=8cm,natwidth=632,natheight=562]{dataexchange.png}\hspace*{\fill}
	\caption{Data exchange between LBM and level set method}
	\label{dataexchange}
\end{figure}
Finally, we can couple the two methods in the following way:
\begin{itemize}
	\item[1.] Create the initial interface $\Gamma_0$
	\item[2.] Run the level set code once to create the surface description for LBM
	\item[3.] Run the LBM using the data from the level set for a prescribed number of steps
	\item[4.] Pass the current velocity field to the level set code and run it for the same number of steps (see Fig. \ref{dataexchange})
	\item[5.] Repeat step 3 and 4 until the end of the simulation.
\end{itemize}

@@

\bibitem{GRcg2} {\sc A.K. Gunstensen, D.H. Rothman}, {\em Macroscopic modeling of immiscible flows in three dimensions by a lattice Boltzmann method}, Europhys. Lett. 18 (2) (1998) 157-161

@@

\bibitem{Koerner} {\sc C. K\"orner, M. Thies, T. Hofmann, N. Th\"urey, U. R\"ude}, {\em Lattice Boltzmann Model for Free Surface Flow for Modeling Foaming}, J. Stat. Phys. 121 (2005) 179-196

@@

\bibitem{LBM3} {\sc S. Succi}, {\em The Lattice Boltzmann Equation for Fluid Dynamics and Beyond}, Clarendon Press, Oxford, ISBN 978-0-19-850398-9 (2001)

@@

\bibitem{Krueger} {\sc T. Kr\"uger, F. Varnik, D. Raabe}, {\em Shear stress in lattice Boltzmann simulations}, Phys. Rev. E 79 (4) (2009) 046704
